A Data-Driven Framework for Neural Field Modeling

### 2011-01-13 ltx/first_order_framework/response_to_reviews.tex

    \begin{enumerate}
        \item Do not use real data for the validation of their approach

	\emph{The authors appreciate that the only way to truly validate the proposed framework is with real data. However, before validating the method on real data we believe that it is essential to first validate the estimation method using accepted models and synthetic data, where the true and estimated parameters can be compared. In an effort to make the method employed in the paper more explicit, we have added ``To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison between the estimated and true parameters'' to the introduction. Another paper demonstrating the application of the method with data recorded from a Utah array implanted in an epilepsy surgical patient is currently in preparation. We have also added that validating the framework on real data is noted as future work in the conclusion.}

        \item Assume that a few kernel basis functions accurately represent the actual connectivity kernel, which might not be the case when using real data;

	\emph{The examples presented in the paper use the Mexican hat connectivity kernel. The framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions. The mathematical formulation does not place any further restrictions on the form of the spatial mixing kernel. We have added statements to the methods section \parham{(after equation~13)} to be more explicit in describing subtleties of the method.}

@@

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.

\emph{The relationship to our basis function decomposition to Galerkin's projection has been added to the derivation of the state-space model. }

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?

\emph{Fourier basis functions would be a correct alternative to the case examined in this paper. In fact Fourier basis functions or any other orthonormal basis functions would simplify the state-space representation of the model, since $\Gamma$ (defined in equation (20)) of the manuscript would be the identity matrix. However, we have chosen Gaussian basis functions due to their semi-compact support. \parham{A comment has been added for possible alternative basis functions under equation (23) of the manuscript.}}

### 2011-01-13 ltx/first_order_framework/response_to_reviews.tex

        \item Assume that a few kernel basis functions accurately represent the actual connectivity kernel, which might not be the case when using real data;

	\emph{The examples presented in the paper use the Mexican hat connectivity kernel. The framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions. The mathematical formulation does not place any further restrictions on the form of the spatial mixing kernel. We have added statements to the methods section \parham{(after equation~13)} to be more explicit in describing subtleties of the method.}

        \item Do not consider the EEG/MEG lead field when constructing their mapping between the membrane voltage and the electrophysiological data (observation function)

	\emph{In response to this important point, the authors have made an effort to be clear that we are not considering scalp EEG to be a valid data source for the estimation framework. Amendments have been made to the Introduction (last paragraph), Method \parham{(after equation 9)}, and Discussion (Implications for Experimental Design). We have primarily developed the framework for higher density intracranial measurements. This is explained in the spatial frequency analysis method and discussion.}

@@

    \begin{enumerate}
        \item The discrepancy of the results in Experiment II where the weights of the basis functions are not in agreement with the actual parameters while the reconstructed kernels seem satisfying.

		\emph{These results show that the system may not be identifiable in non-ideal, practical situations. Nevertheless, the net shape of the estimated kernels will be in accordance to actual data. \parham{These results indicate that the solution to the system may not be unique in non-ideal, practical situations. Nevertheless, the system is identifiable in terms of the net shape of the estimated kernel.}}

        \item The use of Euler discretisation and how this affects the system's behavior.

		\emph{\parham{Amendments have been made to Appendix A, where the discretization is performed. We have provided a proof for the stability of the discrete-time system. In addition,  We have added reference for the further analysis of the error introduced by Euler discretization method.}}

@@

    \section{Reviewer 2}

    We thank the reviewer for their encouraging review. Each point raised is dealt with in turn.

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.

	\emph{\parham{To address this point, we have added an extra paragrapgh to 'Neural Field Model' section of the Methods. There we discuss model mismatch and what we think is a good way to deal with it.}}

@@

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.

\emph{The relationship to our basis function decomposition to Galerkin's projection has been added to the derivation of the state-space model. }

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?

\emph{Fourier basis functions would be a correct alternative to the case examined in this paper. In fact Fourier basis functions or any other orthonormal basis functions would simplify the state-space representation of the model, since $\Gamma$ \parham{(defined in equation (20))} of the manuscript would be the identity matrix. However, we have chosen Gaussian basis functions due to their semi-compact support. \parham{A comment has been added for possible alternative basis functions under equation (23) of the manuscript.}}

\item In equation 29 \parham{(now equation (35))}, what would a reasonable $\boldsymbol{\nu}_c$ be for a real neuronal system (mammalian) given the number of neurons and scale of electrodes? I am concerned that it might be much higher than any of us wish to consider. One would perhaps be saved by the filter characteristics of brain (e.g. Phys. Rev. E 73, 051911 (2006)).

@@

\emph{The mean spatial frequency content is actually lower than one might first expect. We have added `From the work of (Freeman et al, 2000), the expected bandwidth of the human neocortex is approximately 0.1-0.4 cycles/mm which requires a sensor spacing of approximately 1.25 mm to avoid distortions due to aliasing' to the Frequency Analysis section of the paper to clear this up. We have also added to the discussion. Note, we expect the spatial bandwidth of the cortex to be governed by the connectivity kernel, since it is equivalent to a spatial low-pass filter acting on the firing rates.}

\item The two covariance inflation parameters in equation 41 (\parham{now equation (47)}) make me cringe. How are we best to adjust these fudge factors?

\emph{Values for the two parameters in equation 41 \parham{(now equation (47))} along with the relevant references ,that describe the rationale for parameter choice, have been added to the manuscript.}

### 2011-01-13 ltx/first_order_framework/LaTeX_Manuscript.tex
editted the greenie bastard

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
\parham{The post-synaptic membrane potential is in fact the temporal convolution between the synaptic kernel and the average firing rate
\begin{equation}\label{FullConvolutionModel}
 v\left(\mathbf r,t\right)= \left(h\ast g\right)\left(\mathbf r,t\right)
\end{equation}
where $\ast$ denotes the convolution operator.}
\dean{(reviewer 2, comment 3:) \parham{To obtain (To arrive at)} the standard integro-differential equation form of the model, we use the fact that the synaptic response kernel is a Green's function of a linear differential equation defined by the differential operator $\textrm{D}=d/dt + \zeta$. A Green's function satisfies
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation} 
where $\delta(t)$ is the Dirac-delta function. Applying the differential operator $\textrm{D}$ \parham{to equation~\ref{FullConvolutionModel}} gives
\begin{align}
 \textrm{D}v\left(\mathbf r,t\right)&= \textrm{D}\left(h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\textrm{D}h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\delta \left(t\right)\ast g\right)\left(\mathbf r,t\right)\\
&=g\left(\mathbf r,t\right)
\end{align}
} This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretize time using a first-order Euler method (see \ref{App:Time Discretization} for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean Gaussian process with spatial covariance function $\gamma(\mathbf{r}-\mathbf{r}')$ (\cite{Rasmussen2005}). The disturbance is added to account for model uncertainty and unmodeled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

>>>>

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
To obtain (To arrive at) the standard integro-differential equation form of the model, we use the fact that the synaptic response kernel is a Green's function of a linear differential equation defined by the differential operator $\textrm{D}=d/dt + \zeta$. A Green's function satisfies
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation} 
where $\delta(t)$ is the Dirac-delta function. Applying the differential operator $\textrm{D}$ \parham{to equation~\ref{SpikesToPotential}} gives
\begin{align}
 \textrm{D}v\left(\mathbf r,t\right)&= \textrm{D}\left(h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\textrm{D}h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\delta \left(t\right)\ast g\right)\left(\mathbf r,t\right)\\
&=g\left(\mathbf r,t\right)
\end{align}
This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretize time using a first-order Euler method (see \ref{App:Time Discretization} for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean Gaussian process with spatial covariance function $\gamma(\mathbf{r}-\mathbf{r}')$ (\cite{Rasmussen2005}). The disturbance is added to account for model uncertainty and unmodeled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.
